---
title: "Seasonal data plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seasonal data plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(seasoner)
library(ggplot2)
```

We use the `respnet` data in this package, which includes weekly hospitalization rates for COVID-19, influenza, and RSV from the [RESP-NET](https://www.cdc.gov/resp-net/dashboard/index.html) surveillance system. First, visualize the data through time, coloring by network.

```{r}
seasoner::respnet |>
  ggplot(aes(week_ending, rate, color = network)) +
  geom_line() +
  theme_classic()
```

Then, align dates to the 2024/2025 season. Color by season, and facet by network.

```{r}
seasoner::respnet |>
  dplyr::mutate(
    week_ending_align = align_weeks(week_ending, 2024),
    season = date_to_season(week_ending),
    season_label = paste0(season, "/", season + 1)
  ) |>
  ggplot(aes(week_ending_align, rate, color = season_label)) +
  facet_grid(rows = vars(network), scales = "free_y") +
  geom_point() +
  geom_line() +
  theme_classic()
```

Note that, as discussed in the aligning weeks vignette, that 2022/2023 and 2023/2024, when aligned to 2024/2025 have a first date that aligns to a day actually before the start of the 2024/2025 season. Visually, this is not particularly distracting.
